Associative Data Structures for Website Filtering

### 2014-04-17 paper/paper.tex
implementation errors, typos

\title{Associative Data Structures for Website Filtering}
\author{Pieter Brederode, Timo Koppenberg, \\ Jelco Bodewes, Anton Golov}

@@

    \begin{abstract}
        In this paper, we will research associative data structures. This is done in the context of
        a website filter. We look at three different data structures, the hash map, binary search tree
        and the skip list. We insert keys of three different types, namely IP adress, domain name and
        full path URL. We judge the competency of our data strucutre based on insert time, query time
        and memory usage, inserting data sets of sizes $2^x$ for $8 \le x \le 20$. We describe the details
        of our research and the implementation using C++. The results are that hash maps are much faster
        than the other structures, where skip lists use the least memory and binary search trees are
        only useful with very few insertions of simple keys and more insertions than queries.
    \end{abstract}

    \section{Introduction}
    A website filter is a program that catagorises websites in order to restrict access to designated
    sites a user might visit. Commonly, the filter includes a blacklist of sites a user may not visit
    unless they have certain credentials, or alternatively, includes a whitelist so it can blacklist
    everything else. As the world-wide web is growing, these lists have grown larger to a point where
    the speed of the underlying data structure has become a key issue. The time of both inserting and
    looking up websites must be kept minimal. This all must be done without too much of a trade-off
    between speed and memory usage.

@@

    In the following paper, we will attempt to answer these questions. The next section will give details
    about our research plan and the scopes of our research. The section thereafter is about the implementation
    of our setup in C++. Next, we will state our results, after which we will discuss set results and provide
    several conclusions.    
    %We did research about what data structure could best be used to represent a blacklist in a 
    %web-filter application.  We compared the insertion speed, query speed, and memory usage of three
    %data structures: the binary search tree, the hash table, and the skip list.  We did this because
    %we really liked these three structures. 
    \section{Implementation}
    \subsubsection*{Datastructures}

@@

    \subsection{Possible errors caused by the implementation being not entirely perfect}

    We really did our best to make sure these kinds of things didn't happen.  However, there were
    still a number of measurements that do not follow from the underlying theory and are likely a
    consequence of the hardware involved.

    In particular, constructing a small hash table after constructing many big ones takes an
    absurdly long time.  Only the first hash table is affected, and the time is comparable to the
    time required to construct a 4k element hash table, which is just silly.  It probably has
    something to do with the allocator but the reason for such behaviour is still unclear.  The fact
    a hashtable needs contiguous memory probably has something to do with it, as well as the
    allocator returning memory to the operating system.  Still, the latter should also have happened
    with the other structures.  Bizarre.

    A similar issue occurs around 4k or 8k (depending on data structure).  In all three cases, the
    cost of insertion and querying grows at a higher rate than previously.  This is difficult to
    quantify\footnote{Can we do some sort of linear regression to prove this is the case?}, but it
    seems like some extra costs are added around then.  We suspect it is related either to the
    allocator having to get more memory or the CPU cache running out and more cache misses occuring.
    We'd measure that too if we had more time/resources.

    A possibly related issue is that the standard deviation of our measurements rises around that
    point.  That strengthens the possibility of it being allocator-related, as that would involve a
    system call and could cause scheduling issues (despite the system being otherwise very close to
    idle).

    Also, we have counterintuitive results that show that inserting or querying a full path is
    faster than querying only the domain name.  This can partially be explained\footnote{Not really,
    but ehhh.} by the fact that the string implementation we used was based on reference counting.
    (If it'd also provide SSO, this would be easy to explain, but the darn thing doesn't.)

    \section{Conclusions}
    Our main point of research was researching which associative data structure performed best on
    a various number of insertions, in terms of ths speed of insertions and look-ups, and memory.
    In terms of speed we can state with certainty that the hash table is the superior data structure.
    It outperforms the other two in almost every category, most significantly in larger datastructures where
    it is often takes less than half as much time as the skiplist and less than twothirds the time of the binary search tree.
    In most cases, the binary search tree is a bit slower, where the skip list is the slowest structure in
    virtually every case. The only motivation for using a hashmap for this problem appears to be in memory usage,
    although the advantage gained from this is quite small.

    Our secondary point of research encompassed the effect of key complexity on associative data structures.
    Our different types of keys were IP adresses, domain names and full path URLs. We have seen that as expected,
    IP adresses as keys give a significant speedup over the other two, and use less memory as well. Interesting to
    note is that we found the only advantage of the binary search tree over the hashmap here, in that insertions of
    a small amount of IP adresses, up to around $2^13$ insertions, is faster. So when writing an application with very
    few keys of low complexity, with more insertions than queries, the binary search tree may be the best option. Lastly,
    we have found that the domain key type does not outperform the full path key type. It is not significantly better
    if it is better at all.
    %The skip list uses the least memory but is awful in every other imaginable way.  The hashmap is
    %usually the best, except when we're dealing with a small number of integer keys and doing far more
    %inserts than queries.  Note that this doesn't actually happen in practice.

    \section{Reflection}
    In general, we are glad with the results of our research. We were able to provide clear and concise answers to our
    research questions. Even though our research questions may appear broad, we have defined a good scope in our research
    proposal and have virtually not mended this scope in our research. One small change is that we also included datasets of
    size $2^8$ because we saw some interesting results in very small sets. In our implementation we have changed quite a few
    things because we ran into pratical problems. Some of these required us to change the allocator and generate our keys
    in a slightly different fashion, which we all documented in this paper.

    Since we have found that the hash map is the best associative data structure overal for this problem, we might do further
    research in different kind of hash maps. We might try different hash functions, different methods of collision resolution,
    and perhaps survey different kinds of concurrent hash maps.

>>>>

\title{Associative Data Structures for Website Filtering}
\author{Jelco Bodewes, Pieter Brederode \\ Anton Golov, Timo Koppenberg }

@@

    \begin{abstract}
        In this paper, we will research associative data structures for searching. This is done in the context of
        a website filter. We look at three different data structures: the hash map, the binary search tree
        and the skip list. We insert keys of three different types, namely IP adress, domain name and
        full path URL. We judge the competency of our data strucutre based on insertion times, query times
        and memory usage, inserting data sets of sizes $2^x$ for $8 \le x \le 20$. We describe the details
        of our research and the implementation using C++. The results are that hash maps are much faster
        than the other structures, skip lists use the least memory and binary search trees are
        only useful with very few insertions of simple keys and more insertions than queries.
    \end{abstract}

    \section{Introduction}
    A website filter is a program that categorises websites in order to restrict access to designated
    sites a user might visit. Commonly, the filter includes a blacklist of sites a user may not visit
    unless they have certain credentials, or alternatively, includes a whitelist so it can blacklist
    everything else. As the world-wide web is growing, these lists have grown larger to a point where
    the speed of the underlying data structure has become a key issue. The time of both inserting and
    looking up websites must be kept minimal. This all must be done without too much of a trade-off
    between speed and memory usage.

@@

    In the following paper, we will attempt to answer these questions. The next section will provide details
    about our research plan and the scopes of our research. The section thereafter is about the implementation
    of our setup in C++. Next, we will state our results, after which we will discuss set results and provide
    several conclusions.    
    %We did research about what data structure could best be used to represent a blacklist in a 
    %web-filter application.  We compared the insertion speed, query speed, and memory usage of three
    %data structures: the binary search tree, the hash table, and the skip list.  We did this because
    %we really liked these three structures. 
    \section{Implementation}
    \subsubsection*{Datastructures}

@@

    \subsection{Errors possibly caused by an imperfect implementation}

    There were a number of measurements that did not follow the underlying theory and are likely a
    consequence of the hardware involved.

    In particular, constructing a small hash table\footnote{state the number of elements in a "small" hash table}
    after constructing many big ones takes a disproportionally long time.  Only the first hash table is affected, and the time is 			comparable to the time required to construct a hash table with 4000 elements, which is illogical.
    We assume this is caused by the allocator but the reason for such behaviour is still unclear.  The fact
    that a hashtable needs contiguous memory is likely to have some influence, as well as the
    allocator returning memory to the operating system.  However, we reason the latter should also have happened
    with the other structures.

    A similar issue occurs around 4000 or 8000 elements (depending on data structure).  For all three
    datastructures, the cost of inserting and querying grows at a higher rate than before.  This is difficult to
    quantify\footnote{Can we do some sort of linear regression to prove this is the case?}, but it
    appears some extra load is added at that point.  We suspect it is related either to the
    allocator having to acquire more memory or the CPU cache running out and more cache misses occuring.
    We would measure this as well, if there was more time.

    A potentially related issue is that the standard deviation of our measurements rises around the
    same point.  That increases the likelihood of it being allocator-related, as that involves a
    system call and could cause scheduling issues\footnote{elaborate/ref?}, despite the system being 
    otherwise very close to idle.

    Also, we acquired counterintuitive results showing that inserting or querying a full path is
    faster than querying only the domain name.  This can partially be explained\footnote{Not really,
    but ehhh.}by the fact that the string implementation we used was based on reference counting.
    %(If it'd also provide SSO, this would be easy to explain, but the darn thing doesn't.)

    \section{Conclusions}
    Our main point of research was researching which associative data structure performed best on
    a various number of insertions, in terms of the speed of insertions and look-ups, and memory usage.
    In terms of speed we can state with certainty that the hash table is the superior data structure.
    It outperforms the other two in almost every category, most significantly in larger datastructures where
    it often takes less than half as much time as the skiplist and less than two thirds the time of the binary search tree.
    In most cases, the binary search tree is a bit slower, whereas the skip list is the slowest structure in
    virtually every case. The only motivation for using a hashmap for this problem appears to be in memory usage,
    although the advantage gained from this is relatively small.

    Our secondary point of research encompassed the effect of key complexity on associative data structures.
    Our different types of keys were IP adresses, domain names and full path URLs. We have seen that as expected,
    IP adresses as keys provide a significant improvement over the other two, and uses less memory as well. Interesting to
    note is that we found the only advantage of the binary search tree over the hashmap here, is that insertions of
    a small amount of IP adresses, up to around $2^13$ insertions, is faster. So when writing an application with very
    few keys of low complexity, and with more insertions than queries, the binary search tree may be the best option. Finally,
    we have found that the domain key type does not outperform the full path key type. It is not significantly better
    if it is better at all.
    %The skip list uses the least memory but is awful in every other imaginable way.  The hashmap is
    %usually the best, except when we're dealing with a small number of integer keys and doing far more
    %inserts than queries.  Note that this doesn't actually happen in practice.

    \section{Reflection}
    In general, we are content with the results of our research. We were able to provide clear and concise answers to our
    research questions. Even though our research questions may appear broad, we have defined a good scope in our research
    proposal and have virtually not altered this scope in our research. One small change is that we also included datasets of
    size $2^8$ because we saw some interesting results in very small sets. In our implementation we have made some significant
    changes because we ran into pratical problems. Some of these required us to change the allocator and generate our keys
    in a slightly different fashion, which we all documented in this paper.

    Since we have found that the hash map is the best associative data structure overal for this problem, we could do further
    research in different kind of hash maps. We might try different hash functions, different methods of collision resolution,
    and perhaps survey different kinds of concurrent hash maps.
